\documentclass{article}

\usepackage{sty/final}

\nipsfinalcopy

\title{Faster Unsupervised Morphology Induction}

\author{
Victor Chahuneau\\
\texttt{vchahune@cs.cmu.edu}
\And
Peter Schulam\\
\texttt{pschulam@cs.cmu.edu}
\And
Phani Gadde\\
\texttt{pgadde@cs.cmu.edu} \\
}

\begin{document}

\maketitle

\begin{abstract}
  Morphology studies the rules that govern the ways that the morphemes
  of a language may be put together in meaningful ways. For example,
  inflecting the word \textit{walk} to form its present participle
  form \textit{walking} indicates that the action is ongoing in the
  sentence \textit{I am walking}. In many languages, morphology can
  play an even larger role, and may encode many inflectional
  categories such as tense, aspect, mood, number, gender, or case. The
  syntax and semantics of a sentence are highly dependent on this
  information, and it is therefore crucial that natural language
  processing systems use morphological analyses.  In this work, we
  address the issue of efficiency in Bayesian nonparametric models of
  morphology. Specifically, we formulate and implement a model that
  extends the seminal work of \cite{goldwater2011}. Our new model is
  inspired by a recent reparameterization of DP mixture models that
  allows the Chinese Restaurant Process (CRP) to be split among
  several compute nodes in order to speed up inference
  \cite{williamson2013}. In Section \ref{sec:existing-models}, we
  review the morphological model that we have extended and discuss
  some of the linguistic assumptions encoded in the model. With a
  thorough understanding of the baseline in place, we discuss the DP
  mixture model in Section \ref{sec:dpmm} and review the recent work
  on parallelizing inference. Additionally, we draw connections
  between the baseline model in Section \ref{sec:existing-models} and
  DP mixture models, which motivates the design of our parallelized
  model. In Section \ref{sec:parallel-goldwater} we introduce our
  primary contribution and discuss the decisions made when designing
  and implementing our proposed model. Section \ref{sec:evaluation}
  describes the datasets on which we evaluated our model, and presents
  results demonstrating that our parallel model successfully recovers
  the same analyses as the baseline and significantly improves the
  speed of inference when compared against a baseline serial model.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

\input{tex/01-introduction}

\section{Existing Models for Unsupervised Morphology}
\label{sec:existing-models}

\input{tex/02-existing-models}

\section{Dirichlet Process Mixture Models}
\label{sec:dpmm}

\input{tex/03-dpmm}

\section{Parallelized Goldwater Model}
\label{sec:parallel-goldwater}

\input{tex/04-parallel-goldwater}

\section{Evaluation}
\label{sec:evaluation}

\input{tex/05-evaluation}

\section{Conclusion}
\label{sec:conclusion}

\input{tex/06-conclusion}

\bibliographystyle{plain}
\bibliography{bib/final}

\end{document}
